Simulation of Tsunami impact upon Coastline

### 2016-04-07 paper/paper.tex
Equation typo fix.
Equation typo fix 2.
Equation typo fix 3.
Revision after review.

\section{Introduction}
Simulations of natural phenomena are a precious tool for analysis and understanding of the
processes behind them as well as the implications of those. Especially fluid dynamics is
one of the fields most benefitted by the explosion of high performance parallel computing
architectures of the last years. Tsunamis are one of the most devastating natural
disasters, with much attention drawn to them lately, especially after the 2004 Indian
Ocean tsunami and the 2011 T\={o}hoku earthquake and tsunami, two of the largest incidents
of modern history. Multiscale modelling of tsunami generation, propagation and impact is a
trending research area as respective simulations give valuable insights into the
underlying mechanisms and relations between the various stages of an unfolding tsunami
incident, while also facilitating the assessment of potential hazard it poses upon impact
on a coastline.

A tsunami is a series of waves in a water body caused by an impulsive disturbance that
vertically displaces a large volume of water. Tsunamis are generated by earthquakes,
volcanic eruptions, landslides and other such events which have the potential to transmit
a huge amount of mechanical energy to a water column. At large scales, tsunami propagation
in the ocean is usually modelled through various versions of the Shallow Water Equations,
derived from the Navier-Stokes equations in the case where the horizontal length scale is
much greater than the vertical one. Conversely, simulations of tsunami impact is usually
carried out using other methods, since complex dynamics must be taken into account. One of
the most used methods for simulating complex flows with multiple boundary interactions is
Smoothed Particle Hydrodynamics, initially proposed in the 1970s for the treatment of
compressible flows in astrophysical problems. Since then, it has been applied to a wide
variety of fields such as aerodynamics, geology, computer graphics and engineering with
exceptional results. In our simulation, emphasis was given to the conservation of momentum
and its distribution upon the coastline during the tsunami impact. Therefore, we adopted
SPH as our method of choice due to the unparalleled advantages it offers, relating to its
properties as lagrangian method.

\section{Related Work}
Computational fluid dynamics is a very active area of research, as it relates to a wide
spectrum of applications in computer graphics, engineering and science. The SPH method was
developed independently by Gingold and Monaghan \cite{gingold1977375} and Lucy
\cite{lucy19771013} in 1977. Since then it has been employed in a wide variety of problems
and applications, proving itself to be a flexible and attractive method for the simulation
of complex, multicomponent/multiphase flows. Its flexibility is shown by the numerous
adaptations and customizations it has undergone to suit many diverse problem domains.

Relating to the method itself, Desbrun and Gascuel \cite{desbrun1996smoothed} used it to
animate highly deformable bodies of various stiffness and viscosity, while proposing
important extensions like the ``spiky'' pressure smoothing kernel and discussing
implementation issues such as the surface reconstruction from density isosurfaces. Later,
M\"{u}ller et al. \cite{muller2003particle} used this work as a basis for fluid simulation
in interactive applications. Becker and Teschner \cite{becker2007weakly} developed Weakly
Compressible SPH, where the ideal gas equation of state is replaced with the much more
strict Tait equation to reduce compressibility to a user-defined upper bound, thus
avoiding an inefficient explicit Poisson equation solver. In recent years, Solenthaler and
Pajarola \cite{solenthaler2009predictive} improved WCSPH by proposing
Predictive-Corrective Incompressible SPH, where a prediction-correction scheme is used to
compute particle pressures through iterative density constraint satisfaction and
propagation through the fluid, until the final tolerance conditions are met. Implicit
Incompressible SPH was then proposed by Ihmsen et al. \cite{ihmsen2014implicit}, in which
the density is predicted using a discretized form of the pressure Poisson equation, as
obtained from the combination of a direct discretization of the continuity equation and
symmetric pressure forces.

Due to its adaptability and generality, SPH has been the method of choice for many
applications and relative extensions. Macklin and M\"{u}ller \cite{macklin2013position}
enriched SPH with positional geometric constraints, thus enforcing incompressibility while
maintaining stability and allowing for large timesteps. Rigid-fluid interaction is one of
the strong aspects of the method, as is shown by Akinci et al. \cite{akinci2012versatile},
where rigid surfaces are sampled by boundary particles, which adaptively contribute to
fluid properties to address boundary region deficiencies and inhomogeneities, in order to
incorporate rigid bodies to a unified hydrodynamic framework. Debroux et
al. \cite{debroux2001three} used SPH for the simulation of tsunami impact on real
coastline topography, where important features like the energy and speed of the wave were
measured over the course of time. Finally, a thorough overview of various implementation
algorithms and data structures together with their advantages and drawbacks is provided by
Dom\'{i}nguez et al. \cite{dominguez2011}.

\section{Proposed framework}
Our proposed simulation framework is based on SPH, a lagrangian fluid simulation method
whose core notion is the discretization of the fluid into particles, which serve as
interpolation points for the estimation of fluid properties in space. Advantages of this
method include the exact treatment of advection, the natural way of dealing with special
interface interactions, the inherent conservation of significant quantities (mass,
momentum, energy) and the self-adaptivity of computational load to the fluid location and
state in the flow domain. To efficiently ensure compressibility in degenerate cases and
undersampled boundary regions while maintaining relatively large timesteps, SPH was
enhanced with explicit solving of geometric constraints between particles. These
constraints are solved by the Bullet physics engine, within which fluid particles are
represented as rigid bodies, while also being subject to SPH forces as computed at each
timestep of the simulation.

\subsection{SPH method}
Starting from the identity:
\begin{equation*}
  f(\vec{r}) = \int_Vf(\vec{x}) \delta(\vec{r} - \vec{x}) d\vec{x},
\end{equation*}
where $\delta(\vec{r})$ the Dirac delta function and $\vec{x} \in V$, one can obtain a
more general interpolation rule by substituting $\delta(\vec{r})$ with a smoothing kernel
$W(\vec{r}, h)$:
\begin{equation*}
  \label{eq:c-approx}
  f(\vec{r}) \approx \int_V f(\vec{x}) W(\vec{r}-\vec{x}, h) d\vec{x}
\end{equation*}
whose limit when $h\to0$ approaches the delta fucntion and is normalized to unity:
\begin{equation*}
  \label{eq:kernel-properties}
  \lim_{h\to0}W(\vec{r}, h) = \delta(\vec{r})\ \ \
  \text{and}\ \ \
  \int_VW(\vec{r}, h) d\vec{x} = 1
\end{equation*}
The smoothing radius $h$ serves as a cutoff radius in the smoothing process, as particles
beyond that distance have no contribution to the sum, i.e. $W(r,h) = 0$ when $r>h$. For
the discrete case, where $f$ is discretized to particles with density $\rho$ and mass $m$,
the weighting ratio $m/\rho$ can be used to construct a weighted sum interpolant for any
field $A$:
\begin{equation}
  \label{eq:d-approx}
  A(\vec{r}) = \sum_i \frac{m_i}{\rho_i} A(\vec{r}_i) W(\vec{r}-\vec{x}_i, h)
\end{equation}
which lies at the heart of SPH formulation. According to this, the gradient can be
computed by the following approximation:
\begin{equation}
  \label{eq:d-grad}
  \nabla A(\vec{r}) = \sum_i \frac{m_i}{\rho_i} A(\vec{r}_i) \nabla W(\vec{r} - \vec{x}_i, h)
\end{equation}
The obvious advantage of this is the exclusive dependence on the smoothing kernel
gradient, which can be precomputed for sensible kernel choices. However, this formula can
lead to unsymmetric pair forces, compromising the conservation of linear and angular
momentum of the system. To symmetrize these forces depending on gradients (like those
originating from pressure differences), we can use the product rule:
\begin{equation*}
  \label{eq:grad-identity}
  \nabla \left( \frac{P}{\rho} \right) =
  \frac{\nabla P}{\rho}-
  \frac{P}{\rho^2} \nabla \rho
  \hspace{10pt} \Leftrightarrow \hspace{10pt}
  \nabla P = \rho \left[ \frac{P}{\rho^2} \nabla \rho + \nabla \left( \frac{P}{\rho}
    \right) \right]
\end{equation*}
to obtain an alternative approximation of gradient
\begin{align}
  \label{eq:grad-est}
  \nabla P & = \rho \left[ \frac{P}{\rho^2} \sum_i \frac{m_i}{\rho_i} \rho \nabla
             W(\vec{r}-\vec{r}_i, h)
             +
             \sum_i \frac{m_i}{\rho_i} \frac{P_i}{\rho_i} \nabla W(\vec{r}-\vec{r}_i, h)
             \right] \nonumber \\
           & = \rho \sum_i m_i \left(\frac{P}{\rho^2} + \frac{P_i}{\rho_i} \right)
             \nabla W(\vec{r}-\vec{r}_i, h)
\end{align}
which is antisymmetric for all interacting particle pairs. Viscosity forces on the other
hand, are proportional to the laplacian of the velocity field:
\begin{equation}
  \label{eq:lapl-est}
  \nabla^2\vec{v} = \sum_i \frac{m_i}{\rho_i} (\vec{v}_i - \vec{v}) \nabla^2
  W(\vec{r}-\vec{r}_i, h)
\end{equation}
and are always antisymmetric, since they depend on velocity difference
$\vec{v}_i - \vec{v}$ between particles.  In each timestep of the simulation, the density
of all particles is first computed according to equation (\ref{eq:d-grad}), as it depends
only on the relative position of those. The pressure at each particle location is then
obtained from its respective density through an equation of state. Subsequently, the
pressure and viscosity forces are computed from the particle data and integrated back into
the position and velocity of the particles.

\subsection{Implementation}
Each simulation under our implementation consists of two elements, the coastline terrain
(static) and the tsunami wave (dynamic). For the simulation setup, terrain is imported
from a suitable 3D geometry definition file format, the initial conditions of the
impacting wave (position and velocity) are configured, and the desired discretization
resolution for the fluid is set. From these conditions, the terrain and fluid are
initialized, and key parameters are computed as subsequently described. Terrain geometry
is scaled by a user-supplied factor and docked to the origin of coordinates, while fluid
particles are initially placed into an Hexagonal Close-Packed lattice, to achieve the
densest possible packing and symmetry:
\begin{equation}
  \label{eq:hcp}
  [x, y, z] = \left[ 2i+[(j+k) \bmod 2], \sqrt{3}[j + \frac{1}{3} (k \bmod 2)],
    \frac{2\sqrt{6}}{3} k \right] r
\end{equation}
In this configuration, the smoothing radius $h$ is computed such that each particle has
approximately 50 neighbours (following the empirical rule established in the
literature). The timestep is determined according to the Courant-Friedrichs-Lewy
criterion:
\begin{equation}
  \label{eq:cfl}
  \delta t_{\text{CFL}} = C \frac {\delta x}{v_{\text{max}}},
\end{equation}
for values of Courant number $C\approx0.5$ with characteristic length $\delta x$ equal to
the effective particle radius and maximum velocity $v_{\text{max}}$ determined by the
maximum potential energy in the initial configuration. Following standard practice, we
used three different smoothing kernels for density, pressure gradient and velocity
laplacian computation:
\begin{equation}
  \label{eq:poly6}
  W_{\text{poly6}}(r, h) = \frac{315}{64 \pi h^9}
  \begin{cases}
    (h^2 - r^2)^3 & 0 \leq r \leq h\\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
\begin{equation}
  \label{eq:spiky}
  W_{\text{spiky}}(r, h) = \frac{15}{\pi h^6}
  \begin{cases}
    (h - r)^3 & 0 \leq r \leq h\\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
\begin{equation*}
  \label{eq:spiky-gradient}
  \nabla W_{\text{spiky}}(r, h) = \frac{-45}{\pi h^6} (h-r)^2
\end{equation*}
\begin{equation}
  \label{eq:viscosity}
  W_{\text{viscosity}}(r, h) = \frac{15}{2 \pi h^3}
  \begin{cases}
    - \frac{r^3}{2h^3} + \frac{r^2}{h^2} + \frac{h}{2r} - 1 & 0 \leq r \leq h\\
    0 & \text{otherwise},
  \end{cases}\\
\end{equation}
\begin{equation*}
  \label{eq:viscosity-laplacian}
  \nabla^2 W_{\text{viscosity}}(r, h) = \frac{45}{\pi h^6} (h-r)
\end{equation*}

For the computation of the pressure from the estimated fluid density, the ideal gas
equation of state is used:
\begin{equation}
  \label{eq:ideal-state}
  P = k(\rho - \rho_0),
\end{equation}
according to which the pressure is proportional to the difference of the current from the
rest density. The major problem with this equation of state are the compressibility issues
that have been shown to exist in simulations using it. A frequently proposed solution is
to replace the above with the Tait equation of state:
\begin{equation}
  \label{eq:tait-state}
  P = B \left( \left( \frac{\rho}{\rho_0} \right)^\gamma - 1 \right),
\end{equation}
where usually $\gamma=7$ and $B$ a proportionality constant controlling the tolerance to
density fluctuations. This equation is much more punishing on density fluctuations away
from the rest density, therefore requiring significantly smaller timesteps to ensure
stability. Here we follow a different approach, inspired by Position Based Dynamics, in
which particles are represented by spherical rigid bodies. The most important advantage of
this technique is the elegant handling of boundary, undersampled and degenerate fluid
regions that tend to arise very frequently in simulations of free flows. The adoption of
this approach allows to treat boundary collisions in a simple manner, while naturally
enforcing incompressibility near boundaries and free surfaces. In these regions,
estimators fail to describe the actual flow regime due to neighbour particle
shortage. This undersampling creates the need for correction procedures, usually involving
ghost/boundary particles, to avoid pressure instabilities, particle clustering and other
artifacts. On the contrary, no such method is necessary under our representation, where
degenerative cases are handled through geometric constraints thus selectively preventing
unreliable estimations from affecting the flow.

@@

The Bullet physics engine is used to handle the rigid body dynamics. The simulation domain
is divided by a regular grid of spacing equal to the smoothing radius into cells
containing the fluid particles, which are generated as Bullet objects and stored in a
custom cell list data structure consisting of three vectors. As shown in Figure
\ref{fig:lp-grid}, particles are accessed by following pointers through those vectors in
order, with the first encoding the 3D to linear locality preserving mapping of cells on
the second, and the second pointing at first of the cell's particles which are
continuously stored on the third, a dynamic sliding vector. This data structure allows for
quick neighbour search, interaction scanning, exploitation of access patterns of the SPH
algorithm and fast, in-place update. Simulation is following the Bullet framework, with
the SPH code being embedded as an internal timestep tick callback function. We took
advantage of the Bullet infrastructure to extract detailed information about the
collisions between fluid and terrain regarding the resulting impulse, time, and
location. Impulse and particle data are then written to multiple VTK files per
frame. Samples of the smoothed color field (common name in the literature for the field
having the value 1 at particle locations and 0 everywhere else) on the regular cell
lattice are also exported, which are then used to reconstruct the fluid surface as an
isosurface of that field. At the end of the simulation a cumulative impulse heatmap along
with the scaled and docked terrain model are also provided.

\section{Results}
Multiple simulations were carried out using different models of urban coastline, in order
to gain a significant and diverse dataset of impulses exerted over the duration of the
impact. Tsunamis are vastly different from the usual wind-induced sea waves in that they
have far longer wavelength and carry much greater total energy, appearing as a rapidly
rising tide instead of a breaking waves. Accounting for these facts, we chose to represent
the tsunami wave as a water volume invading the coastline with an initial velocity.

\begin{figure}[h!]
  \includegraphics[width=\textwidth]{figures/paraview.png}
  \caption{Example visualization of impact data in Paraview\textsuperscript{TM}. Fluid
    surface is reconstructed as isosurface of fluid density, while impulses on the
    selected region are vi\-su\-a\-li\-zed as points on the 3D terrain model and in a
    histogram grouped by their magnitude.}
  \label{fig:visualization}
\end{figure}

>>>>

\section{Introduction}

\paragraph{} Simulations of natural phenomena are a precious tool for analysis and
understanding of the processes behind them as well as the implications of
those. Especially fluid dynamics is one of the fields most benefitted by the explosive
growth of high performance parallel computing architectures of the last years. Tsunamis
are one of the most devastating natural disasters, with much attention drawn to them
lately, especially after the 2004 Indian Ocean tsunami and the 2011 T\={o}hoku earthquake
and tsunami, two of the largest incidents in modern history. Multiscale modelling of
tsunami generation, propagation and impact is a trending research area as respective
simulations give valuable insights into the underlying mechanisms and relations between
the various stages of an unfolding tsunami incident, while also facilitating the
assessment of potential hazard it poses upon impact on a coastline.

\paragraph{} A tsunami is a series of waves in a water body caused by an impulsive
disturbance that vertically displaces a large volume of water. Tsunamis are generated by
earthquakes, volcanic eruptions, landslides and other such events which have the potential
to transmit a huge amount of mechanical energy to an overlying or adjacent water
volume. On a macroscopic level, tsunami propagation across the ocean as well as coastline
inundation and runup are usually simulated through methods based on various versions of
the shallow water equations, which are derived from the Navier-Stokes equations if the
horizontal length scale is much greater than the vertical one. Conversely, the
aforementioned methods are not appropriate in smaller scale simulations of the impact upon
the coastline, since in order to obtain a reliable estimation of the forces exerted on the
terrain, complex dynamics of fluid-structure interactions have to be accounted for. One of
the most used methods for simulating complex flows with multiple boundary interactions is
Smoothed Particle Hydrodynamics, initially proposed in the 1970s for the treatment of
compressible flows in astrophysical problems. The method has enjoyed extensive adoption,
having been applied to numerous fields ranging from aerodynamics and geology to
engineering and computer graphics with exceptional results, lending itself well to
adaptations and extensions. The main contribution of our approach is the emphasis given to
the explicit conservation of momentum and the detailed data collection and visualization
relating to its distribution upon the coastline during the tsunami impact. Towards this
end, we adopted SPH enriched with geometric constraints as our method of choice due to the
unparalleled advantages it offers, directly linked to its properties as a lagrangian
method.

\section{Related Work}

\paragraph{} Computational fluid dynamics is a very active area of research, as it relates
to a wide spectrum of applications in computer graphics, engineering and science. The SPH
method was developed independently by Gingold and Monaghan \cite{gingold1977375} and Lucy
\cite{lucy19771013} in 1977. Since then, it has been employed in a wide variety of
problems and applications, proving itself to be a flexible and attractive method for the
simulation of complex, multicomponent/multiphase flows. Its flexibility is shown by the
numerous adaptations and customizations it has undergone to suit many diverse problem
domains.

\paragraph{} Tsunami simulations have been shown to be a valuable tool in the literature,
aiding in estimation of key parameters of a tsunami incident, such as its propagation
course and coastal inundation. Simulations of the devastating 2004 Indian Ocean tsunami
have been carried out by Wang and Liu \cite{wang2007} and Ioualalen et
al. \cite{ioualalen2007}, using nonlinear shallow water equations and nonlinear Boussinesq
models respectively. In both cases, good agreement between calculations and field
measurements of inundation and runup indicate the robustness of these approches. Samaras
et al. \cite{samaras2015} used a 2D horizontal model based on the higher oder Boussinesq
equations to simulate tsunami-induced coastal inundation in two relevant areas of interest
in the Mediterranean. Using two different models, Kakinuma \cite{kakinuma2008} reproduced
a past incident with sufficient accuracy. The governing equations for both were the
continuity and Reynolds equations for incompressible fluids in porous media, while water
surface displacement was determined by the vertically integrated continuity equation for
the first and the 3D Volume of Fluid method for the second model.

\paragraph{} Relating to the SPH method, Desbrun and Gascuel \cite{desbrun1996smoothed}
used it to animate highly deformable bodies of various stiffness and viscosity, while
proposing important extensions like the ``spiky'' pressure smoothing kernel and discussing
implementation issues such as the fluid surface reconstruction from density
isosurfaces. Later, M\"{u}ller et al. \cite{muller2003particle} used this work as a basis
for fluid simulation in interactive applications. Becker and Teschner
\cite{becker2007weakly} developed Weakly Compressible SPH, where the ideal gas equation of
state is replaced with the much more strict Tait equation to reduce compressibility to a
user-defined upper bound, thus avoiding an inefficient explicit Poisson equation
solver. In recent years, Solenthaler and Pajarola \cite{solenthaler2009predictive}
improved WCSPH by proposing Predictive-Corrective Incompressible SPH, where a
prediction-correction scheme is used to compute particle pressures through iterative
density constraint satisfaction and propagation through the fluid, until the final
tolerance conditions are met. Implicit Incompressible SPH was then proposed by Ihmsen et
al. \cite{ihmsen2014implicit}, in which the density is predicted using a discretized form
of the pressure Poisson equation, as obtained from the combination of a direct
discretization of the continuity equation and symmetric pressure forces.

\paragraph{} SPH has been employed in many simulations involving wave-coastline
interactions. Debroux et al. \cite{debroux2001three} used SPH for the simulation of
tsunami impact on real coastline topography, where important features like the energy and
speed of the wave were measured over the course of time. G\'{o}mez-Gesteira and Dalrymple
\cite{gomez2004} simulated the impact of a wave generated by a dam break against a tall
structure with a 3D SPH method, showing good agreement with lab measurements of velocities
and forces. Dalrymple and Rogers \cite{dalrymple2006} gave a comprehensive overview of
techniques regarding usage of SPH for simulation of water waves and their interaction with
structures, while also highlighting its suitability for close-up examinations of small
regions. Many studies have considerably improved upon this line of research, yielding
exceptional results. For example, St-Germain et al. \cite{st-germain2014} used a 3D WCSPH
model to examine and analyze spatially and temporally forces exerted by tsunami-like
hydraulic bore upon freestanding structures.

\paragraph{} Due to its generality and adaptability, SPH has been the method of choice for
many applications and relative extensions. Macklin and M\"{u}ller
\cite{macklin2013position} enriched SPH with positional geometric constraints, thus
enforcing incompressibility while maintaining stability and allowing for large
timesteps. Rigid-fluid interaction is one of the strong aspects of the method, as is shown
by Akinci et al. \cite{akinci2012versatile}, where rigid surfaces are sampled by boundary
particles, which adaptively contribute to fluid properties to address boundary region
deficiencies and inhomogeneities, in order to incorporate rigid bodies to a unified
hydrodynamic framework. Finally, a thorough overview of various implementation algorithms
and data structures together with their advantages and drawbacks is provided by
Dom\'{i}nguez et al. \cite{dominguez2011}.

\section{Proposed framework}

\paragraph{} Our proposed simulation framework is based on SPH, a lagrangian fluid
simulation method whose core notion is the discretization of the fluid into particles,
which serve as interpolation points for the estimation of fluid properties in
space. Advantages of this method include the exact treatment of advection, the natural way
of dealing with special interface interactions, the inherent conservation of significant
quantities (mass, momentum, energy) and the self-adaptivity of computational load to the
fluid location and state in the flow domain. To efficiently ensure incompressibility in
degenerate cases and undersampled boundary regions while maintaining relatively large
timesteps, SPH was enhanced with explicit solving of geometric constraints between
particles. These constraints are solved by the Bullet physics engine, within which fluid
particles are represented as rigid bodies, while also being subject to SPH forces as
computed at each timestep of the simulation.

\subsection{SPH method}

\paragraph{} Starting from the identity:
\begin{equation*}
  f(\vec{r}) = \int_Vf(\vec{x}) \delta(\vec{r} - \vec{x}) d\vec{x},
\end{equation*}
where $\delta(\vec{r})$ the Dirac delta function and $\vec{x} \in V$, one can obtain a
more general interpolation rule by substituting $\delta(\vec{r})$ with a smoothing kernel
$W(\vec{r}, h)$:
\begin{equation*}
  \label{eq:c-approx}
  f(\vec{r}) \approx \int_V f(\vec{x}) W(\vec{r}-\vec{x}, h) d\vec{x}
\end{equation*}
whose limit when $h\to0$ approaches the delta fucntion and is normalized to unity:
\begin{equation*}
  \label{eq:kernel-properties}
  \lim_{h\to0}W(\vec{r}, h) = \delta(\vec{r})\ \ \
  \text{and}\ \ \
  \int_VW(\vec{r}, h) d\vec{x} = 1
\end{equation*}
The smoothing radius $h$ serves as a cutoff radius in the smoothing process, as particles
beyond that distance have no contribution to the sum, i.e. $W(r,h) = 0$ when $r>h$. For
the discrete case, where $f$ is discretized to particles with density $\rho$ and mass $m$,
the weighting ratio $m/\rho$ can be used to construct a weighted sum interpolant for any
field $A$:
\begin{equation}
  \label{eq:d-approx}
  A(\vec{r}) = \sum_i \frac{m_i}{\rho_i} A(\vec{r}_i) W(\vec{r}-\vec{r}_i, h)
\end{equation}
which lies at the heart of SPH formulation. According to this, the gradient can be
computed by the following approximation:
\begin{equation}
  \label{eq:d-grad}
  \nabla A(\vec{r}) = \sum_i \frac{m_i}{\rho_i} A(\vec{r}_i) \nabla W(\vec{r} - \vec{r}_i, h)
\end{equation}
The obvious advantage of this is the exclusive dependence on the smoothing kernel
gradient, which can be precomputed for sensible kernel choices. However, this formula can
lead to unsymmetric pair forces, compromising the conservation of linear and angular
momentum of the system. To symmetrize these forces depending on gradients (like those
originating from pressure differences), we can use the product rule:
\begin{equation*}
  \label{eq:grad-identity}
  \nabla \left( \frac{P}{\rho} \right) =
  \frac{\nabla P}{\rho}-
  \frac{P}{\rho^2} \nabla \rho
  \hspace{10pt} \Leftrightarrow \hspace{10pt}
  \nabla P = \rho \left[ \frac{P}{\rho^2} \nabla \rho + \nabla \left( \frac{P}{\rho}
    \right) \right]
\end{equation*}
to obtain an alternative approximation of gradient
\begin{align}
  \label{eq:grad-est}
  \nabla P & = \rho \left[ \frac{P}{\rho^2} \sum_i \frac{m_i}{\rho_i} \rho_i \nabla
             W(\vec{r}-\vec{r}_i, h)
             +
             \sum_i \frac{m_i}{\rho_i} \frac{P_i}{\rho_i} \nabla W(\vec{r}-\vec{r}_i, h)
             \right] \nonumber \\
           & = \rho \sum_i m_i \left(\frac{P}{\rho^2} + \frac{P_i}{\rho_i^2} \right)
             \nabla W(\vec{r}-\vec{r}_i, h)
\end{align}
which is antisymmetric for all interacting particle pairs. Viscosity forces on the other
hand, are proportional to the laplacian of the velocity field:
\begin{equation}
  \label{eq:lapl-est}
  \nabla^2\vec{v} = \sum_i \frac{m_i}{\rho_i} (\vec{v}_i - \vec{v}) \nabla^2
  W(\vec{r}-\vec{r}_i, h)
\end{equation}
and are always antisymmetric, since they depend on velocity difference
$\vec{v}_i - \vec{v}$ between particles.  In each timestep of the simulation, the density
of all particles is first computed according to equation (\ref{eq:d-grad}), as it depends
only on the relative position of those. The pressure at each particle location is then
obtained from its respective density through an equation of state. After this step, the
pressure and viscosity forces are computed from the particle data and integrated back into
the position and velocity of the particles.

\subsection{Implementation}

\subsubsection{Simulation initialization} Each simulation under our implementation
consists of two elements, the coastline terrain (static) and the tsunami wave
(dynamic). For the simulation setup, terrain is imported from a suitable 3D geometry
definition file format, the initial conditions of the impacting wave (position and
velocity) are configured and the desired discretization resolution for the fluid (i.e. the
number of particles it is discretized into) is set. From these conditions, the terrain and
fluid are initialized and key parameters are computed as subsequently described. Terrain
geometry is uniformly scaled by a user-supplied factor and docked to the origin of
coordinates and the particle effective radius $r$ is determined such that the specified
discretization resolution is achieved. Fluid particles are initially placed on a Hexagonal
Close-Packed lattice covering the fluid initial position, in order to accomplish the
densest possible packing and symmetry:
\begin{equation}
  \label{eq:hcp}
  [x, y, z] = \left[ 2i+[(j+k) \bmod 2], \sqrt{3}[j + \frac{1}{3} (k \bmod 2)],
    \frac{2\sqrt{6}}{3} k \right] r
\end{equation}
In this configuration, the smoothing radius $h$ is computed such that each particle has
approximately 50 neighbours (following the empirical rule established in the
literature). The timestep is calculated according to the Courant-Friedrichs-Lewy
criterion:
\begin{equation}
  \label{eq:cfl}
  \delta t_{\text{CFL}} = C \frac {\delta x}{v_{\text{max}}},
\end{equation}
for values of Courant number $C\approx0.5$ with characteristic length $\delta x$ equal to
the particle effective radius and maximum velocity $v_{\text{max}}$ determined by the
maximum potential energy in the initial configuration.

\subsubsection{SPH parameters} Following standard practice, we used three different
smoothing kernels for density, pressure gradient and velocity laplacian computation,
respectively:
\begin{equation}
  \label{eq:poly6}
  W_{\text{poly6}}(r, h) = \frac{315}{64 \pi h^9}
  \begin{cases}
    (h^2 - r^2)^3 & 0 \leq r \leq h\\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
\begin{equation}
  \label{eq:spiky}
  W_{\text{spiky}}(r, h) = \frac{15}{\pi h^6}
  \begin{cases}
    (h - r)^3 & 0 \leq r \leq h\\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
\begin{equation*}
  \label{eq:spiky-gradient}
  \nabla W_{\text{spiky}}(r, h) = \frac{-45}{\pi h^6} (h-r)^2
\end{equation*}
\begin{equation}
  \label{eq:viscosity}
  W_{\text{viscosity}}(r, h) = \frac{15}{2 \pi h^3}
  \begin{cases}
    - \frac{r^3}{2h^3} + \frac{r^2}{h^2} + \frac{h}{2r} - 1 & 0 \leq r \leq h\\
    0 & \text{otherwise},
  \end{cases}\\
\end{equation}
\begin{equation*}
  \label{eq:viscosity-laplacian}
  \nabla^2 W_{\text{viscosity}}(r, h) = \frac{45}{\pi h^6} (h-r)
\end{equation*}
In each simulation step iteration, $W_{\text{poly6}}$ is used as smoothing kernel for the
fluid density estimation at each particle, based on its neighbour particle
locations. $W_{\text{spiky}}$ and $W_{\text{viscosity}}$ are substituted into equations
(\ref{eq:grad-est}) and (\ref{eq:lapl-est}), in order to compute forces arising from
pressure differences and viscosity, respectively. Fluid pressure is obtained from the
estimated fluid density through the ideal gas equation of state:
\begin{equation}
  \label{eq:ideal-state}
  P = k(\rho - \rho_0),
\end{equation}
according to which the pressure is proportional to the difference of the current from the
rest density. The major problem with this equation of state are the compressibility issues
that have been shown to exist in simulations using it. A frequently proposed solution is
to replace the above with the Tait equation of state:
\begin{equation}
  \label{eq:tait-state}
  P = B \left( \left( \frac{\rho}{\rho_0} \right)^\gamma - 1 \right),
\end{equation}
where usually $\gamma=7$ and $B$ is a proportionality constant controlling the tolerance
for density fluctuations. This equation is much more punishing on density fluctuations
away from the rest density, therefore requiring significantly smaller timesteps to ensure
stability.

\subsubsection{Enhancements to SPH} In our framework we opted for a different solution,
inspired by Position Based Fluids \cite{macklin2013position}, in which particles are
represented by spherical rigid bodies. The most important advantage of this technique is
the elegant handling of boundary, undersampled and degenerate fluid regions that tend to
arise very frequently in simulations of free flows. The adoption of this approach allows
to treat boundary collisions in a simple manner, while naturally enforcing
incompressibility near boundaries and free surfaces. In these regions, estimators fail to
describe the actual flow regime due to neighbour particle shortage. This undersampling
creates the need for correction procedures, usually involving ghost/boundary particles, to
avoid pressure instabilities, particle clustering and other artifacts. On the contrary, no
such method is necessary under our representation, where degenerative cases are handled
through geometric constraints, thus selectively preventing unreliable estimations from
affecting the flow.

@@

\subsubsection{Data organization and export} Rigid body dynamics are handled by the Bullet
physics engine. A custom cell list data structure (LP grid) was implemented in order to
optimize particle dynamic storage and access time. The simulation domain is divided by a
regular 3D grid of spacing equal to the smoothing radius into cubic cells containing the
fluid particles. Particles are generated as Bullet objects and are stored in the LP grid,
which consists of three storage vectors. As shown in Figure \ref{fig:lp-grid}, particles
are accessed by following pointers through those vectors in order. The \texttt{map} vector
contains pointers to \texttt{anchors} and in turn \texttt{anchors} pointers to
\texttt{particles}, which contains the simulation particles. Each cell's particles are
continuously stored in the \texttt{particles} vector. Each element of \texttt{anchors}
represents a cell and contains a pointer to the start of the region that contains that
cell's particles in \texttt{particles}. Since SPH access patterns are heavily reliant on
neighbour searching, increasing the locality in the linear order in which cells are stored
is clearly beneficial to performance, due to better utilization of cache memory. For that
reason, cells are represented spatially sorted along a space-filling curve in
\texttt{anchors} (and consequently, \texttt{particles}), in order to maximize locality
preservation from the 3D simulation domain to the linear storage. To efficiently address
the cells, one extra layer encoding the locality preserving mapping has to be added, in
the form of the first vector \texttt{map}. To access a cell, the three indices $i_x$,
$i_y$, $i_z$ specifying its position in the grid are converted to a linear index $i$ (a
simple way is through the formula $i = i_z * N_y * N_x + i_y * N_x + i_x$, where $N_x$
$N_y$ $N_z$ is the number of cells along the $x$, $y$ and $z$ axis in the grid
respectively). The \texttt{map} vector serves as a lookup table, so that its $i$-th
element contains the pointer to that cell's representation on \texttt{anchors}. LP grid
advantages include cohesive storage, quick neighbour search and interaction scanning,
exploitation of the SPH algorithm access patterns and fast, in-place update.

\paragraph{} Simulation is following the Bullet framework, with the SPH code being
embedded as an internal timestep tick callback function. We took advantage of the Bullet
infrastructure to extract detailed information about the collisions between fluid and
terrain regarding the resulting impulse, time and location. Impulse and particle data are
then written to multiple VTK files per frame. Samples of the smoothed color field (common
name in the literature for the field having the value 1 at particle locations and 0
everywhere else) on the regular cell lattice are also exported, which are then used to
reconstruct the fluid surface as an isosurface of that field. At the end of the simulation
a cumulative impulse heatmap along with the scaled and docked terrain model is also
provided.

\section{Results}

\paragraph{} Multiple simulations were carried out using different models of urban
coastline, in order to gain a significant and diverse dataset of impulses exerted over the
duration of the impact. Tsunamis are vastly different from the usual wind-induced sea
waves in that they have far longer wavelength and carry much greater total energy,
appearing as a rapidly rising tide instead of breaking waves. Accounting for these facts,
we chose to represent the tsunami wave as a water volume invading the coastline with an
initial velocity.

\begin{figure}[h!]
  \includegraphics[width=\textwidth]{figures/paraview.png}
  \caption{Example visualization of impact data in Paraview\textsuperscript{TM}. Fluid
    surface is reconstructed as an isosurface of the color field, while impulses on the
    selected region are vi\-su\-a\-li\-zed as points on the 3D terrain model and in a
    histogram grouped by their magnitude.}
  \label{fig:visualization}
\end{figure}

\begin{figure}[h!]
  \includegraphics[width=.33\textwidth]{figures/impulse-heatmap-0.png}
  \includegraphics[width=.32\textwidth]{figures/impulse-heatmap-1.png}
  \includegraphics[width=.33\textwidth]{figures/impulse-heatmap-2.png}
  \caption{Tsunami impulse heatmaps for three different urban models visualized under
    Paraview\textsuperscript{TM}. Most of the energy of the waves is absorbed by the first
    obstacle in their way.}
  \label{fig:impulse-heatmaps}
\end{figure}
